Pollution, agricultural productivity, and development: Evidence from coal in plants in India\footnote{All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution\_development}

### 2023-07-10 draft/draft.tex
Updating readme
updated draft; labor intensity

\title{Air pollution and agricultural productivity in a developing country\footnote{I thank Ariel Ortiz-Bobea, Marc Bellemare, and David Sungho Park for comments and suggestions that greatly improved the quality of the paper. The usual disclaimer applies. All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution\_development}}
\author{Joshua D. Merfeld\footnote{KDI School of Public Policy and Management and IZA; \href{mailto:merfeld@kdis.ac.kr}{\nolinkurl{merfeld@kdis.ac.kr}}}}
\date{2023-04-12}

\begin{document}
\maketitle
\begin{abstract}
\noindent I document externalities in the agricultural sector of India due to air pollution. Using variation in pollution induced by changes in wind, I show that higher levels of particulate matter lead to decreased agricultural productivity, with changes in productivity of 10 percent being common. Combining wind direction with the rollout of coal plants, results indicate pollution from coal plants has a larger effect on agricultural productivity than other types of pollution. Given that the agricultural sector is a refuge for the poor in many developing countries, these results suggest that the effects of pollution may hit the poorest particularly hard.\\
\strut \\
\textbf{\textit{Keywords}}: pollution, productivity, agriculture, labor, India\\
\textbf{\textit{JEL Codes}}: H40, I15, J22, O13, Q52, Q53
\end{abstract}

@@

There are no doubt positive effects from this increase in power generation; for example, electricity may lead to economic and productivity growth (Dinkelman 2011; Kline and Moretti 2014; Rud 2012; Van de Walle et al. 2017), though electrification alone may not be a sufficient condition (K. Lee, Miguel, and Wolfram 2020). These improvements do not come without downsides, however, especially when it comes to power derived from coal. Pollutants from coal contribute to climate change and the phase out of coal energy is therefore a major climate goal (IEA 2022). Emissions from the production and burning of coal are also directly harmful to health and the environment, including pollutants like sulfur dioxide, nitrogen oxides, and mercury, to name but a few.\footnote{www.eia.gov/energyexplained/coal/coal-and-the-environment.php} While the use of coal had been declining, recent events -- including Russia's invasion of Ukraine -- have led to an increase in the demand for coal.\footnote{www.npr.org/2022/08/15/1117560560/a-rising-demand-for-coal-amidst-war-in-ukraine}

In addition to coal, pollution more generally is increasing in much of the world. Carbon dioxide equivalent per capita has almost double over the last 50 years in India, more than doubled in Indonesia, and almost tripled in China.\footnote{\url{https://edgar.jrc.ec.europa.eu}} Of course, much of this increase is due to the development process itself; as countries develop, their economies transform, with more people moving out of agriculture and into manufacturing and services (Herrendorf, Rogerson, and Valentinyi 2014). Despite this development, agriculture remains a key sector in many developing countries, employing a large proportion of the poor (Dercon 2009). This also means agriculture is a key sector when it comes to poverty reduction (Christiaensen, Demery, and Kuhl 2011). Given the large effects of pollution on health and previously documented impacts on agriculture, this raises a important question: do rising pollution levels in developing countries threaten agricultural productivity?

@@

Identifying the causal effects of pollution is difficult. For starters, the construction of coal plants is likely to be endogenous, though the relationship between construction and economic conditions is not a priori clear. On the one hand, governments may locate new plants in fast-growing areas, while on the other hand, they may decide the exact location of plants based on the political power of local citizens (Kopas et al. 2020). In the case of the data I use here, coal plants are constructed in areas with higher population and higher literacy. In terms of agriculture, higher productivity predicts more coal plants in a given year as well as the construction of a coal plant over the next decade, indicating possible bias when estimating OLS regressions of agricultural productivity on pollution levels.

To identify the effects of air pollution on agricultural productivity, I identify areas of high pollution emission using the location of coal plants. While coal plants may add emissions, I show that they are located in areas with relatively higher levels of pollution, in general. Using these locations as point sources of pollution, I use plausibly exogenous changes in wind direction -- in a similar spirit to Deryugina et al. (2019) -- to measure the effects of pollution on key agricultural outcomes. I first plot the location of coal plants in India by year. Using a village-level shapefile, I identify all villages that are located within 30km of a coal plant at any time during the sample period, which comprises the 2000s to the 2010s, with differences across outcomes depending on data availability. I then calculate the direction from each coal plant to all villages within 30km. I include all areas, regardless of whether the coal plant has opened yet, to allow me to disaggregate the effects of pollution from coal plants with other kinds of pollution later in the paper. Using daily wind data and these spatial variables, I create an identifier separately for every village for whether a given village is located downwind from a coal plant on any given day. I then aggregate this exposure variable based on the desired analysis; for agricultural productivity, for example, I create an exposure variable measuring the total number of exposure days in the five-month agricultural season, focusing only on the monsoon (Kharif) season. More days being located downwind from a high-pollution location leads to substantially higher average concentrations of particulate matter 2.5 (PM 2.5).\footnote{PM 2.5 is defined as particulate matter with a diameter of less than 2.5 micrometers.}

Using wind direction as an instrument for particulate matter, an increase of PM 2.5 by one percent leads to a large decrease in agricultural productivity, at around 0.73 percent. Given within-village changes in exposure, pollution could be responsible for swings of 7-to-14 percent in agricultural productivity across years. The IV estimates indicate that the effect of pollution is more negative than naïve OLS estimates, which is consistent with the evidence on the construction of coal plants; higher levels of pollution are correlated with higher levels of agricultural productivity, a relationship that the IV strategy adjusts downwards. Robustness checks confirm these main results. Randomization inference shows that the estimates are robust to alternative measures of uncertainty, while leads of pollution do not affect current agricultural productivity.

Heterogeneity analysis shows that the negative effects of pollution are larger in villages that experience higher median exposure to pollution and in villages with higher initial yields. Finally, given how many poor farmers rely on rainfall and how weather shocks increase across the globe, I also look at possible multiplicative effects of shocks by interacting the pollution variables with a rainfall variable. Higher exposure along with less rainfall increases the negative effect of either shock individually, meaning these shocks are compounding. Given concerns regarding agricultural productivity and climate change, the rise of pollution in India may be particularly problematic for poor farmers.

I find that these effects are mostly transmitted by reductions in farm labor during narrow windows of the growing season, using labor allocation data from the National Sample Survey (NSS). While there do not appear to be large changes in labor allocation patterns when looking at the season as a whole, there is a clear pattern when we break down exposure to pollution by month: higher levels of pollution at the beginning of the monsoon season (June and July) lead to lower levels of farm labor allocation, but no changes in non-farm labor allocation. This matches up well with a similar exercise with agricultural yields: pollution in June is the only month in which there is a negative effect of pollution on agricultural yield. In tandem, these two results point to effects on labor as one mechanism driving the total effects, though I cannot rule out direct effects of pollution on land productivity with my data.

The final set of analyses tries to disentangle the effects of pollution from the effects of coal pollution, specifically, by using wind direction and the rollout of coal plants across the country. Combining instrumental variables with a differences-in-differences strategy, I show that agricultural productivity has an elasticity of around -0.15 with respect to non-coal pollution. The opening of a coal plant, however, triples the elasticity, to more than -0.45, indicating that coal plant emissions are particularly harmful to agricultural productivity, at least in India. This is consistent with coal emissions being quite different from other forms of pollution; for example, coal emissions contain much more mercury than other forms of pollution.

This paper contributes to several strands of literature. First, we already have evidence of effects of pollution on different forms of labor activity. For example, general levels of pollution decrease productivity of call center works in China (Chang et al. 2019), lead people to perform worse on cognitive functions (Ebenstein, Lavy, and Roth 2016; Wen and Burke 2022), and also lead to lower levels of farm labor productivity in California, driven by changes in ozone concentration (Graff Zivin and Neidell 2012). Also in China, He, Liu, and Salvo (2019) find small but significant negative effects of pollution on manufacturing productivity, while Chen, Oliva, and Zhang (2022) show pollution leads to labor out migration. Perhaps relevant to my labor allocation results, Hanna and Oliva (2015) show that decreases in pollution lead to increases in labor allocation, while I show the same relationship in at least a subset of months of the agricultural season. Several papers are closely related to this one. Aragón and Rud (2016) show that gold mines in Ghana have substantial negative impacts on agricultural productivity in the surrounding areas. Burney and Ramanathan (2014) model the effects of climate change and pollutants on agricultural productivity in India. Finally, Sanders and Barreca (2022) actually find that coal emissions can be good for agriculture in the United States, due to the presence of sulfur, which can increase agricultural productivity. This paper differs in three important ways. First, I explore effects of pollution more generally as well as from coal plants, which are much more ubiquitous than gold mines across the world. Second, I use a different identification strategy from either paper, relying on wind blowing pollution from coal plants. Finally, agriculture in India and the United States can be quite different, with the latter being much more mechanized. This means effects of pollution could differ substantially.

I also contribute to the more general literature on the effects of pollution on health. Pollution has substantial negative effects on health in both developing and developed countries (Arceo, Hanna, and Oliva 2016) and we have known of these negative effects for many years (Brunekreef and Holgate 2002; Kampa and Castanas 2008; Pope III and Dockery 2006). Currie et al. (2014) review the literature on the health effects of pollution during adolescence; we have robust evidence of pollution exposure leading to increases in infant mortality, including in developing countries (Heft-Neal et al. 2018, 2019). Since these effects are often driven by heart and lung disease and asthma (Kampa and Castanas 2008), this my explain the labor allocation effects I identify in this paper.

The rest of the paper is organized as follows. I first discuss data and then methodology, in two separate sections. The methodology section discusses the empirical strategy and validates the wind direction variable using data on particulate matter. I present the main results in \autoref{results} before concluding in \autoref{conclusion}.

@@

To help understand some of the mechanisms driving the relationship between exposure to emissions and agricultural productivity, I also include household survey data. Unfortunately, I am not aware of any publicly available survey data that covers multiple years and has identifiers below the district level. For example, the ARIS-REDS data\footnote{www.ncaer.org/project/additional-rural-incomes-survey-aris-rural-economic-demographic-survey-reds} includes GPS identifiers for one year, but that only allows GPS matching for a total of two years: 1999 and 2007/2008. The DHS, on the other hand, only has state-level identifiers (except for the newest two waves) and is also not a labor survey. Instead, I use the National Sample Survey (NSS) from the Ministry of Statistics and Programme Implementation.

The NSS is a nationally representative survey that was conducted every couple of years until relatively recently. Many rounds of the NSS also included a module on employment that asked about labor allocation over the previous seven days. I use the 61st, 62nd, 64th, 66th, and 68th rounds of the NSS, all of which include the employment module. The publicly available data also includes the date of survey, which allows me to match pollution exposure to the exact day of the survey.

The biggest issue with using the NSS is that it includes only district-level identifiers. The main exposure variables, on the other hand, are defined at a lower level of aggregation, at the village. I choose to aggregate this village-level exposure to the district level as shown in the second panel of Figure \ref{fig:windexample} -- the district is a randomly selected district in the data, Guna district in Madhya Pradesh. I use the area overlap of the villages with the district to define a weighted average of exposure for the district at a given time. I calculate the area of overlap for each village and the district as well as the overall area of the district and use these variables to define weights for individual villages. Any part of the district that does not overlap with one of the villages -- the white area in Figure \ref{fig:windexample} -- does not have any exposure and, as such, receives a value of zero for exposure.

@@

However, even with the fixed effects, this regression likely suffers from endogeneity; places with more pollution may be very different from places with less pollution. It is difficult to sign the direction of the bias a priori, because it depends on how the coal plants are placed. For example, if coal plants are placed in places in need of power -- for example, due to rapid growth -- then the estimates might be upwards biased. On the other hand, if officials purposefully place coal plants in poorer areas for political reasons, then the estimates might be downwards biased.

So what are the characteristics of areas near the construction of new coal plants? Table \ref{tab:plantresultstable} presents four separate regressions, using state-level fixed effects due to the spatial variation in coal plants. The first two columns use 1991 census data -- downloaded from SHRUG -- and the second two columns use 2001 census data, all at the village level. The first column presents results from regressing a dummy for whether there is a coal plant within 30km in 1991 on several census variables. Villages with higher populations and higher literacy in 1991 are more likely to be near a coal plant at the time.

@@

\caption{\label{tab:plantresultstable}Local characteristics and the construction of coal plants}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{1991 census} & \multicolumn{2}{c}{2001 census} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
  & 1991 & 2001 & 2001 & 2011\\
\midrule
Population (log) & 0.009 & 0.007* & 0.013* & 0.005\\
 & (0.006) & (0.004) & (0.006) & (0.004)\\
Area (log) & -0.007* & -0.004 & -0.007* & -0.004\\
 & (0.004) & (0.005) & (0.004) & (0.003)\\
Literacy (prop) & 0.080** & 0.037 & 0.073* & 0.036\\
 & (0.029) & (0.028) & (0.039) & (0.026)\\
Ag. productivity &  &  & 0.014** & 0.028***\\
 &  &  & (0.005) & (0.009)\\
Pollution &  &  & 0.047 & 0.014\\
(log PM 2.5) &  &  & (0.045) & (0.018)\\
Sub-sample & all & no plant & all & no plant\\
\midrule
Observations & 283,021 & 264,265 & 508,784 & 469,004\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item Note: State fixed effects are included in regressions. Standard errors are in parentheses and clustered at the state level. The outcome in the first column is whether the village is within 30km of a coal plant in 1991. The second column is whether a village in 1991 will have a coal plant in 2001, conditional on not having one in 1991. The last two columns are similarly defined, except using 2001 and 2011 as the years. * p<0.10 ** p<0.05 *** p<0.01
\end{tablenotes}
\end{threeparttable}
\end{table}

The second column keeps the 1991 census data but changes the outcome variable to a dummy for whether there \emph{will be} a coal plant constructed within 30km before 2001. This column restricts the sample to only villages that do not already have a coal plant within 30km in 1991. Population and literacy both remain strong predictors of the building of a new coal plant.

The last two columns are particularly relevant to the current results. They present similar results to those in the first two columns, but use the 2001 census data and a dummy for the presence of a coal plant in 2001 and 2011 in columns three and four, respectively. Importantly, I include agricultural productivity in the monsoon season of 2002 -- the earliest year available -- as well as average pollution during the monsoon season for 1998 through 2000 as additional predictors. In 2001, agricultural productivity is a strong predictor for the presence of a coal plant. Moreover, it is also a significant predictor of the construction of a coal plant between 2001 and 2011 (column four). Given that the coefficient is positive, this indicates that coal plants are more likely to be built in areas with higher agricultural productivity. In other words, any selection in the overall location of coal plants would bias against finding negative effects of pollution on agricultural productivity. However, note that this is not the same as looking at wind direction, though comparing the OLS and IV results below leads to the same conclusion regarding selection. Pollution, on the other hand, is not significant but is positive in both columns (and is also logged, just like productivity).

Given the likely endogeneity, I instead turn to a plausibly exogenous source of variation in pollution: wind direction. While the location of a coal plant might be endogenous, the amount of pollution received by a specific village in a given year can vary randomly based on wind direction. By counting the number of days in a given year that a village is downwind from a polluted area and conditioning on village fixed effects and weather (rainfall and temperature), variation across years is plausibly exogenous.

@@

It is worth taking the time to validate the use of wind direction as a measure of exposure to pollution. Consider the data used for the regressions presented in Table \ref{tab:pollutiontable} of the appendix. The outcome variable is particulate matter -- specifically, \(\mathrm{PM 2.5}\), which is particulate matter no larger than 2.5 micrometers in diameter -- which comes from Hammer et al. (2020). \(\mathrm{PM 2.5}\) is one of the harmful byproducts from coal plants, along with sulfur dioxide, different types of nitrogen oxides, and mercury.\footnote{www.epa.gov/airmarkets/power-plants-and-neighboring-communities} This data is available at the monthly level, so in Table \ref{tab:pollutiontable}, I aggregate wind exposure to the monthly level, as well.

Both the particulate matter data and the wind data is at the village level, which means the regressions in Table \ref{tab:pollutiontable} are at the village level. As such, the geographic fixed effects are village fixed effects and the temporal fixed effects are month fixed effects. Since the exposure variable is at the village level and we follow villages over time, the standard errors are also clustered at the village level. In columns two and four, I switch the month fixed effects for district-month fixed effects, which considers only villages in the same district in the same month.

In addition to serving as an example, Table \ref{tab:pollutiontable} also helps validate the use of wind direction as a proxy for exposure to pollution. The first two columns include all available years of data while the last two columns restrict estimation only to the years for which the agricultural productivity regressions are estimated. Across all four columns, the story is the same: when the wind is blowing in the direction of a village from a coal plant within 30 kilometers, estimated \(\mathrm{PM 2.5}\) is substantially higher.

To put context behind pollution numbers, the World Health Organization's updated guidelines are that average annual exposure should not exceed 15 \(\mathrm{\mu g/m^3}\). For the month, an increase at the midpoint of the estimated range is equal to an increase of 0.2 percent of the maximum recommended mean concentration from the WHO. This is just a single day of wind in the direction of a village. Since this includes all villages located within 30km of a coal plant, this is evidence that wind direction can have serious repercussions on the health of hundreds of millions of people in India.

@@

I first present the reduced form results for agricultural productivity. The outcome of all the regressions in this section is the log of agricultural land productivity, which is defined as tons per hectare. The first set of results are in Table \ref{tab:yieldtable}. The unit of analysis is the village-year -- from 2002 to 2013 -- focusing only on the monsoon (kharif) season. The data used to construct the yield measure is calculated using five months of data, so the wind variable is defined for the same five months and is the sum of days in which the wind is pointed towards the village centroid from a coal plant.

@@

I present these results in Figure \ref{fig:randomplot}. The purple density is the resulting distribution of estimated treatment effects when selecting from all villages within the same year. The density of treatment effects is relatively tight around 0, with the left tail only getting slightly beyond -0.05, compared to the most conservative actual estimated treatment effect of -0.73.

>>>>

\title{Air pollution and agricultural productivity in a developing country\footnote{I thank Ariel Ortiz-Bobea, Marc Bellemare, and David Sungho Park for comments and suggestions that greatly improved the quality of the paper. The usual disclaimer applies. All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution\_development}}
\author{Joshua D. Merfeld\footnote{KDI School of Public Policy and Management and IZA; \href{mailto:merfeld@kdis.ac.kr}{\nolinkurl{merfeld@kdis.ac.kr}}}}
\date{2023-07-10}

\begin{document}
\maketitle
\begin{abstract}
\noindent I document negative externalities of air pollution in the Indian agricultural sector. Using variation in pollution induced by changes in wind across years, I show that higher levels of pollution lead to decreased agricultural productivity, with large changes in productivity being common. The negative effects of pollution are larger in areas growing more labor-intensive crops, indicating that the pollution works at least partly through direct effects on labor productivity. Finally, combining wind direction with the rollout of coal plants, results indicate that pollution from coal plants has a larger effect on agricultural productivity than other types of pollution. Given that the agricultural sector is a refuge for the poor in many developing countries, these results suggest that the negative externalities of pollution may hit the poorest particularly hard.\\
\strut \\
\textbf{\textit{Keywords}}: pollution, productivity, agriculture, labor, India\\
\textbf{\textit{JEL Codes}}: H40, I15, J22, O13, Q52, Q53
\end{abstract}

@@

While there are likely positive effects from this increase in power generation -- for example, electricity may lead to economic and productivity growth (Dinkelman 2011; Kline and Moretti 2014; Rud 2012; Van de Walle et al. 2017), though electrification alone may not be a sufficient condition (K. Lee, Miguel, and Wolfram 2020) -- these improvements do not come without downsides, especially when it comes to power derived from coal. Pollutants from coal contribute to climate change and the phase out of coal energy is therefore a major climate goal (IEA 2022). Emissions from the production and burning of coal are also directly harmful to human health and the environment; pollutants like sulfur dioxide, nitrogen oxides, methane, and mercury are particularly harmful.\footnote{www.eia.gov/energyexplained/coal/coal-and-the-environment.php} While the use of coal had been declining, recent events -- including Russia's invasion of Ukraine -- have led to an increase in the demand for coal.\footnote{www.npr.org/2022/08/15/1117560560/a-rising-demand-for-coal-amidst-war-in-ukraine}

In addition to pollution from coal, pollution more generally is also increasing in much of the world. Carbon dioxide equivalent per capita has almost double over the last 50 years in India, more than doubled in Indonesia, and almost tripled in China.\footnote{\url{https://edgar.jrc.ec.europa.eu}} Much of this increase is due to the development process itself -- as countries develop, their economies transform, with more people moving out of agriculture and into manufacturing and services (Herrendorf, Rogerson, and Valentinyi 2014). Despite this development, agriculture remains a key sector in many developing countries, employing a large proportion of the poor (Dercon 2009). This also means agriculture is a key sector when it comes to poverty reduction (Christiaensen, Demery, and Kuhl 2011). Given the large effects of pollution on health and previously documented impacts on agriculture, this raises a important question: do rising pollution levels in developing countries threaten agricultural productivity?

@@

Identifying the causal effects of pollution is difficult. For starters, the construction of coal plants is likely to be endogenous, though the relationship between construction and economic conditions is not a priori clear. On the one hand, governments may locate new plants in fast-growing areas, while on the other hand, they may decide the exact location of plants based on the political power of local citizens (Kopas et al. 2020). In the case of the data I use here, coal plants are constructed in areas with higher population and higher levels of pollution. In terms of agriculture, higher productivity predicts more coal plants in a given year as well as the construction of a coal plant over the next decade, indicating possible bias when estimating effects of pollution on agricultural productivity.

To identify the effects of air pollution on agricultural productivity, I identify areas of high pollution emission using the location of coal plants. While coal plants may add emissions, I show that they are located in areas with relatively higher levels of pollution, in general. Using these locations as point sources of pollution, I use plausibly exogenous changes in wind direction -- in a similar spirit to Deryugina et al. (2019) -- to measure the effects of pollution on key agricultural outcomes. I first plot the location of coal plants in India by year. Using a village-level shapefile, I identify all villages that are located within 30km of a coal plant at any time during the sample period, which comprises the 2000s to the 2010s, with differences across outcomes depending on data availability. I then calculate the direction from each coal plant to all villages within 30km. I include all areas, regardless of whether the coal plant has opened yet, to allow me to disaggregate the effects of pollution from coal plants with other kinds of pollution later in the paper.

Using daily wind data and these spatial variables, I create an identifier separately for every village for whether a given village is located downwind from a coal plant on any given day. I then aggregate this exposure variable based on the desired analysis; for agricultural productivity, for example, I create an exposure variable measuring the total number of exposure days in the five-month agricultural season, focusing only on the monsoon (Kharif) season. More days being located downwind from a high-pollution location leads to substantially higher average concentrations of particulate matter 2.5 (PM 2.5).\footnote{PM 2.5 is defined as particulate matter with a diameter of less than 2.5 micrometers.}

Using wind direction as an instrument for particulate matter, an increase of PM 2.5 by one percent leads to a large decrease in agricultural productivity, with an elasticity of approximately -0.63 in the most conservative specification. The IV estimates indicate that the effect of pollution is more negative than naïve OLS estimates, which is consistent with the evidence on the construction of coal plants; higher levels of pollution are correlated with higher levels of agricultural productivity, a relationship that the IV strategy adjusts downwards. Robustness checks confirm these main results. Randomization inference shows that the estimates are robust to alternative measures of uncertainty, while leads of pollution do not affect current agricultural productivity.

Heterogeneity analysis shows that the negative effects of pollution are larger in villages that experience higher median exposure to pollution and in villages with higher initial yields. Finally, given how many poor farmers rely on rainfall and how weather shocks are on the rise across the globe, I also look at possible multiplicative effects of shocks by interacting the pollution variables with a rainfall variable. Higher exposure along with less rainfall increases the negative effect of either shock individually, meaning these shocks are compounding. Given concerns regarding agricultural productivity and climate change, the rise of pollution in India may be particularly problematic for poor farmers.

A key question is the mechanism: does pollution reduce labor productivity, land productivity, or both? While previous work from the United States indicates that pollution from coal plants, specifically, can lead to higher \emph{land} productivity (Sanders and Barreca 2022), agriculture in India differs greatly from agriculture in the United States. To test for possible effects on labor productivity, I look at differential effects based on labor intensity of the predominant crops in different villages. Specifically, I separate effects in predominantly rice-growing areas from other areas; since rice is very labor intensive, we would expect to see larger negative effects of pollution in rice-growing regions relative to other regions if one mechanism through which pollution affects productivity is direct effects on labor. I show that pollution negatively effects agricultural productivity in non-rice areas but that the effects in rice-growing areas are around 50 percent larger in magnitude. These results indicate that pollution likely directly affects labor productivity, though I am not able to rule out direct effects on land, either, due to a lack of data.

The final set of analyses tries to disentangle the effects of pollution from the effects of coal pollution, specifically, by using wind direction and the rollout of coal plants across the country. Combining instrumental variables with a differences-in-differences strategy, I show that agricultural productivity has an elasticity of around -0.22 with respect to non-coal pollution. The opening of a coal plant, however, doubles the elasticity, to more than -0.44, indicating that coal plant emissions are particularly harmful to agricultural productivity, at least in India. This is consistent with coal emissions being quite different from other forms of pollution; coal emissions contain different types and quantities of harmful substances, meaning effects may differ greatly.

An important caveat to these results is that I cannot disentangle exactly which pollutants drive results. While I use PM 2.5 as a measure of overall exposure to pollution, effects could be driven by any pollutants carried by the wind. This leaves open an important avenue of future research regarding the effects of different types of pollutants on agricultural productivity.

This paper contributes to several strands of literature. We already have evidence of effects of pollution on different forms of labor activity. For example, general levels of pollution decrease productivity of call center works in China (Chang et al. 2019), lead people to perform worse on cognitive functions (Ebenstein, Lavy, and Roth 2016; Wen and Burke 2022), and also lead to lower levels of farm labor productivity in California, driven by changes in ozone concentration (Graff Zivin and Neidell 2012). Also in China, He, Liu, and Salvo (2019) find small but significant negative effects of pollution on manufacturing productivity, while Chen, Oliva, and Zhang (2022) show pollution leads to labor out migration. Relevant to my results, Hanna and Oliva (2015) show that decreases in pollution lead to increases in labor allocation in another developing country.

Several papers are closely related to this one. Aragón and Rud (2016) show that gold mines in Ghana have substantial negative impacts on agricultural productivity in the surrounding areas. Burney and Ramanathan (2014) model the effects of climate change and pollutants on agricultural productivity in India. Finally, Sanders and Barreca (2022) actually find that coal emissions can be good for agriculture in the United States, due to the presence of sulfur, which can increase agricultural productivity. This paper differs from the previous literature in three important ways. First, I explore effects of pollution more generally as well as from coal plants, which are much more ubiquitous than gold mines across the world. Second, I use a different identification strategy from most other papers, relying on wind blowing pollution from coal plants. Finally, agriculture in India and the United States can be quite different, with the latter being much more mechanized. This means effects of pollution could differ substantially depending on the context, a conclusion that the current results support.

@@

To help understand some of the mechanisms driving the relationship between exposure to emissions and agricultural productivity, I also look at the differential effects by the labor intensity of the predominant crop in each area. For this, I use data from (Monfreda, Ramankutty, and Foley 2008), who estimate area harvested for different crops across the global around the year 2000. I identify areas where rice (paddy) is the most commonly grown crop, under the assumption that rice is more labor intensive than other crops (Vollrath 2011; Michler 2020). Michler (2020), for example, shows that rice, on avearge, uses almost 1,800 hours per hectare, while the second-highest crop in his data, cotton, juses just 850.\footnote{He estimates that wheat, another common crop in India, uses just 600 hours per hectare.}

@@

However, even with the fixed effects, this regression likely suffers from endogeneity; places with more pollution may be very different from places with less pollution. It is difficult to sign the direction of the bias a priori, because it depends on the local context and how the coal plants are placed. For example, if coal plants are placed in places in need of power -- for example, due to rapid growth -- then the estimates might be upwards biased. On the other hand, if officials purposefully place coal plants in poorer areas for political reasons, then the estimates might be downwards biased.

What are the characteristics of areas near the construction of new coal plants? Table \ref{tab:plantresultstable} presents four separate regressions, using state-level fixed effects due to the spatial variation in coal plants. The first two columns use 1991 census data -- downloaded from SHRUG -- and the second two columns use 2001 census data, all at the village level. The first column presents results from regressing a dummy for whether there is a coal plant within 30km in 1991 on several census variables. Villages with higher population densities (higher log population and lower log area) are more likely to be near a coal plant at the time. Urban population proportion is not a significant predictor after conditioning on population and area.

@@

\caption{\label{tab:plantresultstable}Local characteristics and the construction of coal plants}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{1991 census} & \multicolumn{2}{c}{2001 census} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
  & 1991 & 2001 & 2001 & 2011\\
\midrule
Population (log) & 0.012*** & 0.009*** & 0.015*** & 0.006***\\
 & (0.001) & (0.000) & (0.000) & \vphantom{1} (0.000)\\
Area (log) & -0.010*** & -0.005*** & -0.010*** & -0.005***\\
 & (0.001) & (0.000) & (0.000) & (0.000)\\
Urban pop (prop) & -0.029 & 0.006 & -0.003 & -0.007\\
 & (0.037) & (0.035) & (0.022) & (0.018)\\
Ag. productivity &  &  & 0.014*** & 0.028***\\
 &  &  & (0.001) & (0.001)\\
Pollution &  &  & 0.051*** & 0.016***\\
(log PM 2.5) &  &  & (0.002) & (0.001)\\
Sub-sample & all & no plant & all & no plant\\
\midrule
Observations & 283,021 & 264,265 & 508,786 & 469,005\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item Note: State fixed effects are included in regressions. Robust standard errors are in parentheses. The outcome in the first column is whether the village is within 30km of a coal plant in 1991. The second column is whether a village in 1991 will have a coal plant in 2001, conditional on not having one in 1991. The last two columns are similarly defined, except using 2001 and 2011 as the years. * p<0.10 ** p<0.05 *** p<0.01
\end{tablenotes}
\end{threeparttable}
\end{table}

The second column keeps the 1991 census data but changes the outcome variable to a dummy for whether there \emph{will be} a coal plant constructed within 30km before 2001. This column restricts the sample to only villages that do not already have a coal plant within 30km in 1991. The pattern of significance remains the same in column two.

The last two columns are particularly relevant to the current results. They present similar results to those in the first two columns, but use the 2001 census data and a dummy for the presence of a coal plant in 2001 and 2011 in columns three and four, respectively. Importantly, I include agricultural productivity in the monsoon season of 2002 -- the earliest year available -- as well as average pollution during the monsoon season for 1998 through 2000 as additional predictors. In 2001, agricultural productivity is a strong predictor for the presence of a coal plant. Moreover, it is also a significant predictor of the construction of a coal plant between 2001 and 2011 (column four). Given that the coefficient is positive, this indicates that coal plants are more likely to be built in areas with higher agricultural productivity. In other words, any selection in the overall location of coal plants would bias against finding negative effects of pollution on agricultural productivity. However, note that this is not the same as looking at wind direction, though comparing the OLS and IV results below leads to the same conclusion regarding selection.

Pollution is also a significant predictor for the presence of a coal plant in 2001 as well as for whether a coal plant will be constructed before 2011. While the coefficient is relatively small, it is actually quite large given that only three percent of villages in column four will get a new plant in the next decade.

Given the likely endogeneity previewed by these summary statistics, I now turn to a plausibly exogenous source of variation in pollution: wind direction. While the location of a coal plant might be endogenous, the amount of pollution received by a specific village in a given year can vary randomly based on wind direction. By counting the number of days in a given year that a village is downwind from a polluted area and conditioning on village fixed effects and weather (rainfall and temperature), variation across years is plausibly exogenous.

@@

It is worth taking the time to validate the use of wind direction as a measure of exposure to pollution. Consider the data used for the regressions presented in Table \ref{tab:yield3ivmainfirststage} of the appendix. The outcome variable is particulate matter -- specifically, \(\mathrm{PM 2.5}\), which is particulate matter no larger than 2.5 micrometers in diameter -- which comes from Hammer et al. (2020). \(\mathrm{PM 2.5}\) is one of the harmful byproducts from coal plants, along with sulfur dioxide, different types of nitrogen oxides, and mercury.\footnote{www.epa.gov/airmarkets/power-plants-and-neighboring-communities} This data is available at the monthly level, but I aggregate variables to the season variable since that is the level of analysis in the main results.

Both the particulate matter data and the wind data is at the village level, which means the regressions in Table \ref{tab:yield3ivmainfirststage} are at the village level. As such, the geographic fixed effects are village fixed effects and the temporal fixed effects are month fixed effects. Since the exposure variable is at the village level and we follow villages over time, the standard errors are also clustered at the village level. In columns two and four, I switch the month fixed effects for district-month fixed effects, which considers only villages in the same district in the same month.

In addition to serving as an example, Table \ref{tab:yield3ivmainfirststage} also helps validate the use of wind direction as a proxy for exposure to pollution. The first two columns include all available years of data while the last two columns restrict estimation only to the years for which the agricultural productivity regressions are estimated. Across all four columns, the story is the same: when the wind is blowing in the direction of a village from a coal plant within 30 kilometers, estimated \(\mathrm{PM 2.5}\) is substantially higher.

\hypertarget{identifying-mechanisms}{%
\subsection{Identifying mechanisms}\label{identifying-mechanisms}}

To identify possible mechanisms, I identify areas that grow predominantly rice, which is a labor-intensive crop. Specifically, we are interested in the following regression:
\begin{gather} 
y_{it} = \alpha_{i} + \gamma_{t} + \omega PM_{it} + \phi rice_{it} + \beta PM_{it}\times rice_{it} + \delta rain_{it} + \psi temp_{it} + \varepsilon_{it},
\end{gather}
where \(PM_{it}\) denotes the effect of pollution in non-rice areas and \(PM_{it}\times rice_{it}\) is the differential effect in rice-growing areas. Here, I create an indicator variable for rice areas that is time invariant, so \(rice_{it}\) drops out of the estimation. I instrument for \(PM_{it}\) and for \(PM_{it}\times rice_{it}\) using wind and wind interacted with the indicator for rice.

@@

I first present the reduced form results for agricultural productivity. The outcome of all regressions in this section is the log of agricultural land productivity, which is defined as tons per hectare. The first set of results are in Table \ref{tab:yieldtable}. The unit of analysis is the village-season -- from 2002 to 2013 -- focusing only on the monsoon (kharif) season. The data used to construct the yield measure is calculated using five months of data, so the wind variable is defined for the same five months and is the sum of days in which the wind is pointed towards the village centroid from a coal plant.

@@

I present these results in Figure \ref{fig:randomplot} of the appendix. The purple density is the resulting distribution of estimated treatment effects when selecting from all villages within the same year. The density of treatment effects is relatively tight around 0, with the left tail only reaching slightly beyond -0.05, compared to the most conservative actual estimated treatment effect of -0.63.

The within-district correlation in outcomes and treatment is apparent when instead randomly selecting from only villages within the same district in the same year. The distribution is still approximately normal but has a mode of almost -0.4. Nonetheless, the actual treatment effect is larger in magnitude than almost all of the bootstrapped replications. The results are apparently robust to alternative specifications of uncertainty.
